---
title: "Statistical_Rethinking00.Rmd"
author: "Robert A. Stevens"
date: "May 4, 2016"
output: html_document
---

*Statistical Rethinking: A Bayesian Course with Examples in R and Stan*

by Richard McElreath

## Table of Contents

0.0 Preface  
1.0 The Golem of Prague  
2.0 Small Worlds and Large Worlds  
3.0 Sampling the Imaginary  
4.0 Linear Models  
5.0 Multivariate Linear Models  
6.0 Overfitting, Regularization, and Information Criteria  
7.0 Interactions  
8.0 Markov Chain Monte Carlo  
9.0 Big Entropy and the Generalized Linear Model  
10.0 Counting and Classification  
11.0 Monsters and Mixtures  
12.0 Multilevel Models  
13.0 Adventures in Covariance  
14.0 Missing Data and Other Opportunities  
15.0 Horoscopes 

# 0.0 Preface  

## 0.1 Audience  

## 0.2 Teaching strategy  

## 0.3 How to use this book  

What the book assumes

https://cran.r-project.org/other-docs.html

Using the code

```{r, comment=NA}
print("All models are wrong, but some are useful.")
```

```{r, comment=NA}
x <- 1:2
x <- x*10
x <- log(x)
x <- sum(x)
x <- exp(x)
x
```

Optional sections

Rethinking:  Think again

Overthinking:  Getting your hands dirty

```{r, comment=NA}
(log(0.01^200))
(200*log(0.01))
```

The command line is the best tool

How you should work

```{r, comment=NA}
# Load the data:
# car braking distances in feet paired with speeds in km/h
# see ?cars for details
data(cars)
# fit a linear regression of distance on speed
m <- lm(dist ~ speed, data = cars)
plot(dist ~ speed, data = cars)
abline(m)
# estimated coefficients from the model
coef(m)
# plot residuals against speed
plot(resid(m) ~ speed, data = cars)
```

## 0.4 Installing the rethinking R package  

```{r, comment=NA}
# only need to install once
#install.packages(c("coda", "mvtnorm", "devtools"))
library(devtools)
#devtools::install_github("rmcelreath/rethinking")
library(rethinking)
```

## 0.5 Acknowledgements  
